---
title: "Spotify Top Hits from 1970 to 2019"
output: github_document
---

## Overview
The following code is for downloading **audio features** of Spotify's "Top Hits of" playlists, specifically from the years 1970 to 2019. For example the playlist's title for the year 1990 it would be "Top Hits of 1990".

It uses the wrapper library [Spotifyr](https://www.rcharlie.com/spotifyr/) to make http requests to the Spotify Web API.

### Code
The file ```test.R``` is an R script for downloading just one playlist, in order to have an idea of the data shape.

The Rmarkdown file ```spotifyHits1970-2019.Rmd``` contains the necessary code for getting the **audio features** of the Top Hits playlists from 1970-2019. Then it exports the data to the file ```Spotify_TopHits_of_1970_2019.csv```.

### Data
The folder ```data``` contains the following data:

- ```Spotify_TopHits_of_1970_2019.csv``` : audio features for the playlists Top Hits of, from 1970 to 2019, retrieved from Spotify Web API.
- ```test_hits_1970.csv``` : audio features for the playlist Tip Hits of 1970, it's the file used for exploring the data provided by the Spotify API.
- ```playlist_features.rds``` : An RDS object for storing all the columns given by the Spotify Web API response, when making the requests for the playlists of 1970-2019.

## Packages
This packages are required to run the code
```{r, message=FALSE}
library(spotifyr)
library(dplyr)
library(readr)
library(DataExplorer)
```

## Downloading the data

### Authentication
The Spotifyr's function ```get_spotify_access_token()``` gets the **Access Token** with the Spotify Client ID and the Spotify Client Secret.

```{r}
access.token <- get_spotify_access_token(
  client_id = Sys.getenv("SPOTIFY_CLIENT_ID"), 
  client_secret = Sys.getenv("SPOTIFY_CLIENT_SECRET")
)
```


### Playlists IDs
The playlists' names are like "Top Hits of YYYY", where YYYY refer to the year. For example for the year 2000 it would be "Top Hits of 2000". 

The years are in the range [1970, 2019], let's make a vector that stores the names.

```{r}
playlists.names <- paste("Top","Hits", "of", as.character(1970:2019)) 
```

In order to get the ID for each playlist, we can use the `search_spotify()` function to retrieve info about the playlists

```{r}
search.response <- lapply(1:length(playlists.names), function(i) {
  search_spotify(q = playlists.names[i], type = "playlist", 
                 authorization = access.token, 
                 limit = 1)
}
)
```

Let's extract the IDs from the list response and then verify that the playlists are the ones created by Spotify

```{r}
# get the IDs for the elements in the list using lapply
search.response.ID <- lapply(1:length(search.response), function(i){
   # the i-th playlist tibble in the list
   aux.playlist.info <- search.response[[i]]
   # making sure it's the playlist created by Spotify
   aux.playlist.info <- aux.playlist.info %>% 
     filter(name == playlists.names[i] & owner.display_name == "Spotify")
   # get the actual playlist id from the 4th element of the i-th tibble 
   aux.playlist.info[[4]]
}
)

# convert the IDs to character
playlists.IDs <- as.character(unlist(search.response.ID))
head(playlists.IDs, 10)
```

### Getting the playlists' audio features using their Spotify IDs
Now, let's use ```lapply()``` again to get the audio features tibbles in a list

```{r}
features.response <- lapply(1:length(playlists.IDs), function(i) {
  get_playlist_audio_features("Spotify", playlists.IDs[i], 
                 authorization = access.token
                 )
}
)
```





Concatenating the tibbles with ```bind_rows```
```{r}
playlists.features <- bind_rows(features.response)
```

## Verifying the downloaded data

### Counting
The total instances in the dataset

```{r}
nrow(playlists.features)
```

The numbers of songs in each playlist

```{r}
# group by playlist and count
playlists.counts <-  playlists.features %>% group_by(playlist_name) %>% summarise(n = n())
print(playlists.counts, n = nrow(playlists.counts))
```
### Removing NA row
We can see that there's one missing value in the ```playlist_name``` column. Let's locate it.

```{r}
index.na <- which(is.na(playlists.features$playlist_name))
print(paste("The row of the missing playlist:",index.na))
print(playlists.features[(index.na-2):(index.na+2), c(2,6:16)])
```

The missing row belongs to the 1997 playlist, but since all the audio features are missing let's remove it.

```{r}
playlists.features <- playlists.features[-index.na,]
paste("Number of rows left:",nrow(playlists.features))
paste("missing values in playlist_name column:" , 
      sum(is.na(playlists.features$playlist_name)))
```
## Selecting the audio features columns in the data

### Backup
Making a backup for the tibble in an ```.rda```file, which can be later imported into an R session with ```readRDS()``` function.

```{r}
rds.output <- "./data/playlist_features.rds"

if (file.exists(rds.output)) {
   print(paste("The file already exists:", rds.output))
} else {
   saveRDS(playlists.features, file = rds.output)
}
```

### Columns

Now, there may be more missing values in other columns, but before checking that, let's select the columns that contain the audio features fisrt.

It's important to mention that not all the columns are for audio features, some are for IDs or metadata like links to the playlist's cover images that could be useful for a dashboard. 

The columns that will be extracted are

```{r}
columns.features <- c("playlist_id", "playlist_name", "playlist_img", 
                      "playlist_owner_name", "playlist_owner_id",
                      "danceability", "energy", "key", "loudness", 
                      "mode", "speechiness", "acousticness", "instrumentalness", 
                      "liveness", "valence", "tempo", "track.id",
                      "time_signature", "added_at", "track.artists", 
                      "track.disc_number", "track.duration_ms", "track.explicit",
                      "track.name", "track.popularity",
                      "track.preview_url", "track.type",
                      "track.album.release_date", 
                      "track.album.release_date_precision",
                      "track.album.total_tracks",
                      "track.album.type", "track.album.uri"
                      )
```

Selecting the 32 columns listed above and storing the resulting tibble in ```topHits1970_2019```

```{r}
playlists.features.subset <- playlists.features[ ,columns.features]
dim(playlists.features.subset)
```
### Creating a column for the artists' names
Now, using ```glimpse()``` to see the data types in the ```playlists.features.subset``` tibble 

```{r}
glimpse(playlists.features.subset)
```

The ```track.artists``` column is a list of dataframes, where the artists' names are other information are stored. Let's extract those names and add them to a new column for simplicity in later analysis.

```{r}
topHits1970_2019 <- playlists.features.subset %>% 
   mutate(artists.name = lapply(1:nrow(playlists.features.subset), function(i){
      playlists.features.subset$track.artists[[i]][[3]]
   })
   ) %>% 
   mutate(artists.name = as.character(artists.name))
```

### Missing values
Let's look for missing values one last time before exporting the tibble.

```{r}
plot_missing(topHits1970_2019)
```

The result is that only the column ```track.preview_url``` has missing values, the column contains links which open a popup with a 30 sec preview of the track. As long as the audio features have no missing values then there's no problem with this data.

## Exporting the data to .csv
Finally, let's save the data in a ```.csv``` file.

```{r}
file.output <- "./data/Spotify_TopHits_of_1970_2019.csv"

if (file.exists(file.output)) {
   print(paste("The file already exists:", file.output))
} else {
   write_csv(topHits1970_2019, file = file.output)
}
```

So, the file saved has the columns

```{r}
glimpse(topHits1970_2019)
```


**Note:** The column ```artists.name``` contains strings like ```c("artist 1", "artist 2", "artist 3")``` when there are multiple artists. When there's only one artist it's just a string like ```artist 1```.





